Density-dependent selection and the limits of relative fitness

### 2018-03-17 manuscript.tex
l

In a constant environment, and in the absence of crowding, $r_i$ is a constant ``intrinsic growth rate (for concreteness we will use Eq.~\eqref{eq:canonical} as our point of comparison, but the discrete time case is similar). The selection coefficient $s$ is then simply the difference in intrinsic population growth rates. However, growth cannot continue at a constant rate indefinitely. As population density increases, crowding will cause the Malthusian parameters $r_i$ to decline (e.g. \citealt[pp. 203]{begon_1990}). Since different traits are advantageous in crowded and uncrowded conditions, selection in the presence of crowding can be density-dependent, and we would then need to know the dynamics of $N$ to use Eq.~\eqref{eq:canonical}. In general we cannot simply independently specify the dynamics of $N$, because those ecological dynamics are coupled with the evolutionary dynamics of type frequency \citep{travis_2013}. Thus, in a general crowded population the simple procedure of assigning constant relative fitness values to different types has to be replaced with an ecological description of absolute growth rates. Note that frequency-dependent selection does not raise such a severe problem, because a complete description of selection only requires us to model the type frequencies, not also the ecological variable $N$.

In practice, many population genetics models assume that $N$ has reached some fixed equilibrium value, and each type is assigned a constant relative fitness. Selection is typically assumed to operate through viability (e.g. \citealt[pp. 61]{gillespie_2004}), but other important population ecological assumptions --- such as how density is regulated --- are left unspecified. Beyond being conceptually unsatisfactory \citep{doebeli_2017}, we might also doubt whether this density-independent-$s$, constant-$N$ setup is consistent with the ecology of many populations. Much of the seminal literature on density-dependent selection revolves around the connection between selection in crowded populations and the ability to keep growing at higher densities \citep{macarthur_1967,roughgarden_1979}. The classic example is the logistic model, where the type with the greatest equilibrium population density ``$K$'' excludes the others (Fig.~\ref{fig:Ksel}a). Similarly, the ``$R^*$ rule'', a central tenet of resource competition theory, states that when growth is limited by a single homogeneous consumable resource, the type able to deplete the resource to the lowest equilibrium density $R^*$ (which will often entail a different equilibrium population density) excludes the others \citep{grover_1997}. In these examples, both $N$ and $s$ change during, and as a result of, adaptive sweeps. Density also changes in response to selection in the Lotka-Volterra competition model except in special cases \citep{smouse_1976,mallet_2012}. It would therefore seem that the density-independent-$s$, constant-$N$ setup (Fig.~\ref{fig:Ksel}b) is incompatible with a broad class of population ecological processes driving selection.

@@

However, our understanding of the strong-selection, unstable-$N$ regime is

generally limited to a few prominent examples such as the classical density-dependent selection literature discussed above,

and complex age-structured models \citep[Chap. 4]{charlesworth_1994}.

@@

Most models of density-dependent selection are missing one key element that has important ramifications for the validity of the constant-$s$ approximation: they do not have a ``reproductive excess'' of juveniles that are more fragile than their adult counterparts \citep{turner1968population}. Reproductive excesses allow selection to be concentrated at a juvenile ``bottleneck'', which means that the density of adults can remain constant even if strong selection is occurring on juveniles --- rather different from the classic $K$-selection scenario. Although similar ideas featured prominently in   early debates on the regulation of population density (grounded in experiments on flies; \citealt{nicholson_1954}), much of the reproductive excess literature revolves around Haldane's ``cost of selection'' \citep{haldane_1957}, and is also poorly integrated with population ecology. For instance, \cite{kimura1969natural} simply took constant $N$ as a requirement and then derived variants of the logistic model consistent with this. \cite{nei1971fertility} proposed a model with an explicit representation of reproductive excess, but used an unusual model of competition based on pair-wise interactions which was only defined for at most two different types. As a result, the role of reproductive excesses in justifying the use of constant relative finesses is largely verbal.

@@

We now turn to the issue of how density responds to selection, comparing the variable-density lottery and previous models of selection in density-regulated populations  \citep{prout_1980}. Starting with some of the earliest work on the topic \citep{kostitzin_1939,christiansen_2004}, most previous models exhibit some form of ``$K$-selection'' i.e. selection in crowded populations increases equilibrium density. This idea is intuitively appealing because births balance deaths at equilibrium, and adaptive mutants must overcome the existing balance to expand \citep[pp. 227]{kostitzin_1939}. Yet, as we saw in the previous section, $c$-selection has no effect on population density in the variable-density lottery. To make sense of how $c$-selection fits with previous population growth models, we now revisit MacArthur's general treatment of $K$-selection \citep{macarthur_1967}.

>>>>

In a constant environment, and in the absence of crowding, $r_i$ is a constant ``intrinsic growth rate (for concreteness we will use Eq.~\eqref{eq:canonical} as our point of comparison, but the discrete time case is similar). The selection coefficient $s$ is then simply the difference in intrinsic population growth rates. However, growth cannot continue at a constant rate indefinitely. As population density increases, crowding will cause the Malthusian parameters $r_i$ to decline (e.g. \citealt[pp. 203]{begon_1990}). Selection can then be density-dependent, and we would need to know the dynamics of $N$ to use Eq.~\eqref{eq:canonical}. In general we cannot simply independently specify the dynamics of $N$, because those ecological dynamics are coupled with the evolutionary dynamics of type frequency \citep{nagylaki_1979,travis_2013}. Thus, the simple procedure of assigning constant relative fitness values to different types has to be replaced with an ecological description of absolute growth rates. Note that frequency-dependent selection does not raise such a severe problem, because a complete description of selection only requires us to model the type frequencies, not $N$ as well.

In practice, many population genetic models assume that $N$ has reached some fixed equilibrium value, and each type is assigned a constant relative fitness. Selection is typically assumed to operate through viability (e.g. \citealt[pp. 61]{gillespie_2004}), but other important population ecological assumptions --- such as how density is regulated --- are left unspecified. However, there are reasons to doubt that this setup is ecologically valid. Many absolute fitness models of density-regulated growth exhibit a strong connection between selection and the ability to keep growing at higher densities \citep{fisher_1930,kostitzin_1939,macarthur_1967,roughgarden_1979,christiansen_2004}. This includes simple birth-death \citep{kostitzin_1939} and logistic models (Fig.~\ref{fig:Ksel}a; \citealt{macarthur_1962,kimura1969natural,roughgarden_1979,boyce_1984}), more general functional forms for the absolute fitness penalties of crowding \citep{kimura_1978,charlesworth_1971,lande_2009,nagylaki_1979}, and the ``$R^*$ rule'' of resource competition theory (which states that the type able to deplete a shared, homogeneous, limiting consumable resource to the lowest equilibrium density $R^*$ excludes the others; \citealt{grover_1997}). Density also changes in response to selection (even if only transiently) in the Lotka-Volterra competition model except in special cases \citep{smouse_1976,mallet_2012}. This behavior implies that both $N$ and $s$ change during, and as a result of, adaptive sweeps, and has been studied extensively following Fisher's suggestion that evolution tends to increase density in the long term \citep{fisher_1930, macarthur_1962, roughgarden_1979, lande_2009}. It would therefore seem that the density-independent-$s$, constant-$N$ setup (Fig.~\ref{fig:Ksel}b) is incompatible with a broad class of population ecological processes driving selection.

@@

However, the preceding arguments do not imply that the constant-$s$ approximation \textit{only} applies under weak selection and stable $N$. All of the models of density-dependent selection mentioned above are missing one key element that has important ramifications for the validity of the constant-$s$ approximation: a ``reproductive excess'' of juveniles that are more fragile than their adult counterparts \citep{turner1968population}. In complex age-structured models, dependence on density is assumed to come from a ``critical age-group'', and this age group's density responds to selection in much the same way as found in the above models without age structure \citep[Chap. 4]{charlesworth_1994} (e.g. adaptive evolution proceeds in the direction of increasing equilibrium density in the critical age-group). However, in the constant-$s$ approximation, reproductive excesses allow selection to be concentrated at a juvenile ``bottleneck'', which means that the density of adults can remain constant even if strong selection is occurring on juveniles.

Although similar ideas featured prominently in early debates on the regulation of population density (grounded in experiments on flies; \citealt{nicholson_1954}), much of the reproductive excess literature revolves around Haldane's ``cost of selection'' \citep{haldane_1957}, and is also poorly integrated with population ecology. For instance, \cite{kimura1969natural} simply took constant $N$ as a requirement and then derived variants of the logistic model consistent with this. \cite{nei1971fertility} proposed a model with an explicit representation of reproductive excess, but used an unusual model of competition based on pair-wise interactions which was only defined for at most two different types. As a result, the role of reproductive excesses in justifying the use of constant relative finesses is largely verbal.

This behavior is intuitively appealing because births balance deaths at equilibrium, and adaptive mutants must overcome the existing balance to expand \citep[pp. 227]{kostitzin_1939}.

@@

\citep{charlesworth_1971,kimura_1978,leon_1978,nagylaki_1979}

@@

We now turn to the issue of how density responds to selection, comparing the variable-density lottery to previous studies of density-regulated populations \citep{prout_1980}. As we saw in the previous section, $c$-selection has no effect on population density in the variable-density lottery. To make sense of how $c$-selection fits with previous population growth models, we now revisit MacArthur's general treatment of $K$-selection \citep{macarthur_1967}.
